---
title: "p8105_hw2_yr2560"
author: "Yiman Rui"
date: "2026-10-06"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 1
```{r}
library(tidyverse)
library(readxl)
```

Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).
```{r}
transit = read_csv("~/Desktop/Data Science/Data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

transit
  
```
The cleaned dataset contains 1,868 rows and 19 columns. Each row represents a subway station entrance or exit. The variables include the subway line and station name, station latitude and longitude, the routes served by the station, whether entry is allowed, whether a vending machine is present, entrance type, and ADA compliance.

I cleaned the variable names, selected only the requested variables, and converted entry to TRUE and FALSE. At this point, the data are not completely tidy because route information is spread across route1 through route11, so route number is stored in the column names rather than in its own variable.

Distinct station
```{r}
transit |>
  distinct(station_name, line) |>
  nrow()
```
There are 465 distinct stations.

ADA Compliance station
```{r}
transit |>
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```
There are 84 stations that are ADA compliant. 

Proportion
```{r}
transit |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry, na.rm = TRUE)
  ) |>
  pull(proportion)

```
The proportion of station entrances/exits without vending allow entrance is approximately 0.377.

Reformation of data 
```{r}
transit_reformat = transit |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name)

# number of station serve A
transit_reformat |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()

# number of station serve A that ADA compliant 
transit_reformat |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```
There are 60 stations that serve A and there are 17 of them are ADA compliant. 

## Problem 2

Mr.Trash Wheel
```{r}
mr_trash = read_excel("~/Desktop/Data Science/Data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742",
    na = c("", "NA", "N/A", "-")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )
```

Prof Trash Wheel
```{r}
prof_trash = read_excel("~/Desktop/Data Science/Data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139",
    na = c("", "NA", "N/A", "-", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash Wheel"
  )
```

Gwynnda
```{r}
gwynnda_trash = read_excel("~/Desktop/Data Science/Data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394",
    na = c("", "NA", "N/A", "-")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel =  "Gwynnda"
  )

```

Combine 3 dataset
```{r}
trash_wheel = 
  bind_rows(mr_trash, prof_trash, gwynnda_trash) |>
  relocate(trash_wheel)
```

The combined Trash Wheel dataset contains 1,269 observations. Each observation corresponds to one dumpster collection. Important variables include the Trash Wheel name, dumpster number, month and year, date, trash weight in tons, volume in cubic yards, and counts of several types of trash such as plastic bottles and cigarette butts.

