---
title: "p8105_hw2_yr2560"
author: "Yiman Rui"
date: "2026-10-06"
output: github_document
---

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
```

## Problem 1

```{r}
transit = read_csv("data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv") |>
  janitor::clean_names() |>
  select(
    line,
    station_name,
    station_latitude,
    station_longitude,
    route1:route11,
    entry,
    vending,
    entrance_type,
    ada
  ) |>
  mutate(
    entry = case_match(
      entry,
      "YES" ~ TRUE,
      "NO" ~ FALSE
    )
  )

transit
  
```
The cleaned dataset contains 1,868 rows and 19 columns. Each row represents a subway station entrance or exit. The variables include the subway line and station name, station latitude and longitude, the routes served by the station, whether entry is allowed, whether a vending machine is present, entrance type, and ADA compliance.

I cleaned the variable names, selected only the requested variables, and converted entry to TRUE and FALSE. At this point, the data are not completely tidy because route information is spread across route1 through route11, so route number is stored in the column names rather than in its own variable.

Distinct station
```{r}
transit |>
  distinct(station_name, line) |>
  nrow()
```
There are 465 distinct stations.

ADA Compliance station
```{r}
transit |>
  filter(ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```
There are 84 stations that are ADA compliant. 

Proportion
```{r}
transit |>
  filter(vending == "NO") |>
  summarize(
    proportion = mean(entry, na.rm = TRUE)
  ) |>
  pull(proportion)

```
Among station entrances/exits without vending, approximately 37.7% allow entrance.

Reformation of data 
```{r}
transit_reformat = transit |>
  mutate(
    across(route1:route11, as.character)
  ) |>
  pivot_longer(
    route1:route11,
    names_to = "route_number",
    names_prefix = "route",
    values_to = "route_name"
  ) |>
  drop_na(route_name)

# number of station serve A
transit_reformat |>
  filter(route_name == "A") |>
  distinct(station_name, line) |>
  nrow()

# number of station serve A that ADA compliant 
transit_reformat |>
  filter(route_name == "A", ada == TRUE) |>
  distinct(station_name, line) |>
  nrow()
```

There are 60 stations that serve A, of which 17 are ADA compliant. 

## Problem 2

Mr.Trash Wheel
```{r}
mr_trash = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Mr. Trash Wheel",
    range = "A2:N742",
    na = c("", "NA", "N/A", "-")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    sports_balls = as.integer(round(sports_balls)),
    trash_wheel = "Mr. Trash Wheel"
  )
```

Prof Trash Wheel
```{r}
prof_trash = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Professor Trash Wheel",
    range = "A2:M139",
    na = c("", "NA", "N/A", "-", "dive")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel = "Professor Trash Wheel"
  )
```

Gwynnda
```{r}
gwynnda_trash = read_excel("data/202610 Trash Wheel Collection Data.xlsx",
    sheet = "Gwynnda the Good Wheel of the W",
    range = "A2:L394",
    na = c("", "NA", "N/A", "-")
  ) |>
  janitor::clean_names() |>
  filter(!is.na(dumpster)) |>
  mutate(
    year = as.integer(year),
    plastic_bottles = as.numeric(plastic_bottles),
    trash_wheel =  "Gwynnda"
  )

```

Combine 3 dataset
```{r}
trash_wheel = 
  bind_rows(mr_trash, prof_trash, gwynnda_trash) |>
  relocate(trash_wheel)
```

The combined Trash Wheel dataset contains 1,269 observations. Each observation corresponds to one dumpster collection. Important variables include the Trash Wheel name, dumpster number, month and year, date, trash weight in tons, volume in cubic yards, and counts of several types of trash such as plastic bottles and cigarette butts.

Professor trash wheel total weights
```{r}
prof_trash |>
  summarize(
    total_weight = sum(weight_tons, na.rm = TRUE)
  ) |>
  pull(total_weight)
```

Professor trash wheel collected a total of 294.96 tons of trash.

Gwynnda cigarette butts in June 2022
```{r}
gwynnda_trash |>
  filter(
    year == 2022,
    month == "June"
  ) |>
  summarize(
    total_cigarettes = sum(cigarette_butts, na.rm = TRUE)
  ) |>
  pull(total_cigarettes)
```
Gwynnda collected 18120 cigarette butts in June 2022.

## Problem 3

MCI Baseline data
```{r}
mci_base = 
  read_csv("data/MCI_baseline.csv",
                    skip = 1,
                    na = c(".", "NA")) |>
  janitor::clean_names() |>
  mutate(
    sex = case_match(
      sex,
      0 ~ "Female",
      1 ~ "Male"
    ),
    apoe4 = case_match(
      apoe4,
      0 ~ "non-carrier",
      1 ~ "carrier"
    )
  ) |>
  filter(
    is.na(age_at_onset) | age_at_onset > current_age
  )

mci_base
```

```{r}
n_participants =
  nrow(mci_base)

n_mci =
  mci_base |>
  filter(!is.na(age_at_onset)) |>
  nrow()

mean_base_age =
  mci_base |>
  summarize(
    mean_age = mean(current_age, na.rm = TRUE)
  ) |>
  pull(mean_age)

prop_women_apoe4 =
  mci_base |>
  filter(sex == "Female") |>
  summarize(
    proportion = mean(apoe4 == "carrier")
  ) |>
  pull(proportion)

n_participants
n_mci
mean_base_age
prop_women_apoe4
```
After applying the inclusion criterion, the baseline dataset contains 479 participants. Of these participants, 93 developed MCI during follow-up. The average baseline age was 65.0 years. Among women in the study, 30% were APOE4 carriers.

MCI Amyloid data
```{r}
mci_amyloid =
  read_csv(
    "data/mci_amyloid.csv",
    skip = 1,
    na = c(".", "NA")
  ) |>
  janitor::clean_names() |>
  rename(id = study_id) |>
  pivot_longer(
    baseline:time_8,
    names_to = "visit",
    values_to = "amyloid_42_40"
  ) |>
  mutate(
    time = case_match(
      visit,
      "baseline" ~ 0,
      "time_2" ~ 2,
      "time_4" ~ 4,
      "time_6" ~ 6,
      "time_8" ~ 8
    ),
    amyloid_42_40 = as.numeric(amyloid_42_40)
  ) |>
  select(
    id,
    time,
    amyloid_42_40
  )

mci_amyloid
```
The resulting dataset contains one row for each participant at each study visit. The time variable indicates the number of years since baseline, while amyloid_42_40 contains the amyloid beta 42/40 ratio.

Participants only in one dataset
```{r}
base_only = anti_join(
    mci_base,
    mci_amyloid |>
      distinct(id),
    by = "id"
  )
base_only

amyloid_only =
  anti_join(
    mci_amyloid |>
      distinct(id),
    mci_base,
    by = "id"
  )
amyloid_only

nrow(base_only)
nrow(amyloid_only)

```
There are 8 eligible participants who appear only in the baseline dataset and 16 participant IDs that appear only in the amyloid dataset. Some of the amyloid-only IDs correspond to participants removed from the baseline dataset because they did not meet the the inlucsion criterion, while others do not appear in the baseline demographic data. Therefore, the participant lists in the two source datasets are not identical.

Combine two dataset
```{r}
mci_combined = inner_join(
    mci_base,
    mci_amyloid,
    by = "id"
  )

n_participants =
  mci_combined |>
  distinct(id) |>
  nrow()

dim(mci_combined)

n_participants
```
The combined dataset contains 2355 rows and 8 columns, representing 471 unique participants.

Each participant can contribute up to five biomarker observations corresponding to baseline and years 2, 4, 6, and 8. The final dataset contains participant demographic information, APOE4 carrier status, age at MCI onset when applicable, follow-up time, and amyloid beta 42/40 ratio.

Export final data in csv
```{r}
write_csv(
  mci_combined,
  "data/mci_combined.csv"
)
```

